---
title: "Incarceration Trends in the United States"
author: "Peijie Zheng"
date: "2/24/2022"
output: html_document
---

```{r setup, include = FALSE}
source("../source/analysis.R")
library(knitr)
```

## Overview
This report aims to examine the incarceration trends in the United States from 1990 to 2016. Understanding that the prison system functions "differently" for each race, I want to use data to deepen my understanding of the injustice toward minority groups. Thanks to the incarceration data set provided by the Vera Institute, I am able to critically analyze the statistics and gain insights into the incarceration system. For the analyses, I am focusing specifically on black and white people in relation to the prison system. I realize that racism is deeply rooted in the United States that people of color are discriminated against. Comparing the incarceration rate between these groups exposes inequality and unfair treatment, allowing us to rethink the structure of the prison system. Therefore, the variables I will be working with are the total population and prison population of each race.  

## Summary Information
First off, to get an overview of the prison system, I computed the total population in prison for each year from 1990 to 2016. The data shows that the number of people incarcerated increases each year,  from **610946** in 1990 to **1046289** in 2016 peaking at **1260879** in 2009. This suggests the explosive growth in America's prison population, but also implies that criminal justice is a rising issue. To understand how race plays a role, I calculated the average percentage of black and white people in prison. On average, **39%** of the prison population are black people, whereas only **31%** are white people. These percentages suggest that black people are more likely to be treated less fairly and sent to prison even though their total population in the U.S. is much less than the white population. In addition, the county with the highest number of black and white people in prison is Los Angeles County from 1990 to 2016. However, statistics of the number of white people in this county's prison is **193328**, whereas for black people, this number is **527316**. The discrepancy between these two numbers asserts that racism plays a role in the United States' prison system. 

## Time Trend Chart
```{r, echo = FALSE}
plot(prison_pop_trend)
```
<br /> This is a time chart that illustrates the trend of the prison population between different groups of people from 1990 to 2016. I decided to create this chart because it is important to incorporate groups other than black and white people in the narrative to better understand the prison system. From this chart, black, white, and Latinx people are the major groups targeted for incarceration. The prison population for these groups of people is increasing over the years. This supports the fact that our prison system does not address the issue of criminal justice because we would expect the numbers to go down.

## Comparison Chart
```{r, echo = FALSE}
plot(black_white_prison_percentage)
```
<br /> This chart compares the percentages of black and white people in prison with respect to their total population (15 - 64). I believe this chart is the most significant because although we saw that black and white people have roughly the same number in the prison population, these percentages suggest that black people are more likely to be incarcerated than white people. It emphasizes the idea that numbers don't speak for themselves, and having context of these numbers is crucial for us to understand the truths behind incarceration. Looking at the chart, around **1%** of the total black population from 15 - 64 years old are in prison, whereas this number for the white population is only around **0.2%**! The pattern suggests that black people have always had a higher percentage of being incarcerated than white people.

## Map
```{r, echo = FALSE}
plot(black_prison_pop_map)
```
<br /> This map shows the distribution of the black prison population in the United States across each county. I created this map focusing on the distribution because it allows me to get a better sense of where racism is most deeply rooted in terms of incarceration. From this, we can see that majority of the black people in prison are concentrated in California, Texas, Washington, New York, and Southeast states. Even though the color green means each county has around 1000 - 4000 people in prison, given the number of counties that have black people incarcerated, the disproportionate incarceration of people of color remains an issue.