---
title: "Incarceration Trends in the United States"
author: "Peijie Zheng"
date: "2/24/2022"
output: html_document
---

```{r setup, include = FALSE}
source("../source/analysis.R")
library(knitr)
```

## Overview
This report aims to examine the incarceration trends in the United States from 1990 to 2016. Understanding that the prison systems "works" drastically different for each race, I want to use data to deepen my understanding of injustice toward minority groups. Thanks to the incarceration data set provided by the Vera Institute, it allows me to critically analyze and gain insights into the incarceration system. For the analyses, I am focusing specifically on black and white people in relation to the prison system because I am cognizant that racism is deeply rooted and people of color are heavily discriminated. By comparing the incarceration rate between these groups of people, it directly exposes inequality and unfair treatments, allowing us to rethink the structure of the prison system. Thus, the variables I am working with are the total population and prison population of each race. 


## Summary Information
First off, to get an overview of the prison system, I computed the total population in prison for each year from 1990 to 2016. The data shows that number of people prisoned is increasing each year, going from **610946** in 1990 to **1046289** in 2016 peaked at **1260879** in 2009. This suggests the explosive growth in America's prison population, but also implies that criminal justice is a rising issue. To understand how race plays a role in the prison system, I calculated the average percentage of black and white people in prison. On average, **39%** of the prison population are black people, whereas only **31%** are white people. These percentages suggest that black people are more likely to be treated less fairly and sent to prison even though their total population is so much less than white population. Additionally, the county with the highest number of black and white people in prison is the __Los Angeles County__ from 1990 to 2016. However, the statistics for number of white people in this county's prison is **193328**, whereas for black people, this number is **527316**. The discrepancy between these two number asserts that racism plays a role in the United States' prison system. 



## Time Trend Chart
```{r, echo = FALSE}
plot(prison_pop_trend)
```
<br/> This charts show the trend of prison population of the five races from 1990 to 2016. I decided to create this chart because I believe it is important to incorporate groups other than black and white people to fully understand how the prison system functions. From this chart, we can see that black, white, and latinx people are the major groups targeted to be sent to prison. The prison population for these groups of people is increasing over the years. This supports the fact that our prison system does not address the issue of criminal justice because we would expect the numbers to go down.



## Comparison Chart
```{r, echo = FALSE}
plot(black_white_prison_percentage)
```
<br/> This chart compares the percentages of black and white people in prison with respect to their population (15 - 64). I strong believe this chart is the most significant because although we saw that black and white people have roughly the same number in prison population, these percentages suggest that black people are more likely to be prisoned than white people. This emphasizes the idea that numbers don't speak for themselves and having context of these numbers is crucial for us to understand the truth. Looking at the chart, around **1%** of the total black population from 15 - 64 are in prison, whereas this number for white population is only around **0.2%**! The pattern from this chart suggests that black people have always had a higher percentage of being incarcerated than white people. 




## Map
```{r, echo = FALSE}
plot(black_prison_pop_map)
```
<br/> This map shows the distribution of black prison population in the United States across each county. I created this map focusing on the distribution because it allows me to get a better sense of where racism is most deeply rooted in terms of incarceration. From this, we can see that majority of the black people in prison are concentrated in California, Texas, Washington, New York, and Southeast states. Even though the color green suggests each county has around 1000 - 4000 people in prison, given the number of counties that have black people incarcerated, the disproportionate incarceration of people of color remains an issue.
